Land-use legacies and climate change as a double challenge to oak forest resilience: mismatches of geographical and ecological rear edges

### 2020-06-07 man/ecosystems/ms_ecosystems.Rmd
update results

### Climatic data and drought episodes

The Iberian Peninsula underwent several extreme drought episodes in the last three decades [*e.g.* 1995, 1999, 2005, 2012;  @VicenteSerrano2014]. The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], appearing as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3). A drought event was identified using the SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] (SPEI 12-months scale), following a procedure similar to the one proposed by @Spinoni2015. We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: the *duration* as the number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the absolute SPEI values during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

@@

### Statistical analysis

The severe drought events since 1901 were identified using SPEI-12 and regional climatic data. They were characterized in terms of duration, severity, intensity [see @Spinoni2015]. In a first step we explored temporal trends of EVI and BAI variables. Temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test (Figure S7a). Temporal trends for BAI were assessed using mean BAI chronologies. The impact of drought in greenness and growth was exploring using the $EVI_{sa}$ and the mean RWI site chronologies (Figures S7a). Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim` [@Zang2015]. The non-climatic disturbance impacts on tree-growth were evaluated using the site disturbance chronologies.

@@

## Results
### Time trends in vegetation greenness
The analysis of time trends in greenness showed that 78.9% of the EVI pixels followed a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the study period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 2a). A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was found during this drought event, whereas no changes in greenness in response to the 2012 drought were detected (Figure 2b).

### Analysis of radial-growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high elevation were taller and their growth was significantly faster than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 3). The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 3 and S3a). Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 3), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).

@@

The northern site (SJ) showed two major release events (GC > 50% occurring in more than 50% of trees sampled): the first during the 1940s (the most evident) and the second in 1995-2000 (Figure 4). These periods alternated with periods of suppression. By contrast, the two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.

### Resilience to drought events at the ecosystem and individual-tree levels
*Rs* and *Rt* varied in the same direction whereas *Rc* varied inversely to *Rs* and *Rt*. Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and *Rc*, significant for all oak populations (Figure 5). A similar pattern was found for *Rs* but proved significant only for SJ. Importantly, non-significant patterns resulted when we excluded 1995, except for *Rc* in SJ (Figure S5). The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4).

During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1). The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years. Resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 in both variables (Table S2; Figure 6). Thus, recovery values for greenness were higher for 2005 than for the 2012 drought event. Recovery showed a contrasting pattern for EVI and tree growth.

@@

### Tree-growth response to climate
The response of tree growth to water availability was greater than to temperatures. Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure 7a). Nevertheless, differences appeared between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure 7a). In addition, the spring maximum temperature was the most significant limitation for tree growth only for the southern populations (Figure 7b and 7c), whereas minimum and maximum temperatures of the current September positively influenced tree growth only in the northern population.

>>>>

### Climatic data and drought episodes
The Iberian Peninsula underwent several extreme drought episodes in the last three decades [*e.g.* 1995, 1999, 2005, 2012;  @VicenteSerrano2014]. The 2005 and 2012 drought events have been documented as being among the worst in recent decades for the southern Iberian Peninsula [@Pascoa2017], appearing as extreme drought in our climatic data (Figure S1; Table S3). We focused on these two drought events because they were included in the period having remote-sensing information of high spatial resolution (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize the results for 2005 and 2012 and to evaluate forest resilience to drought over a longer term (see Table S3). A drought event was identified using the SPEI (Standardized Precipitation-Evapotranspiration Index) [@VicenteSerrano2010] (SPEI 12-months scale), following a procedure similar to the one proposed by @Spinoni2015. We used 0.5º grid cells covering Sierra Nevada taken from the Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values fall below that threshold for at least two consecutive months. For each drought event, we computed: the *duration* as the number of consecutive months with the SPEI lower than a certain threshold; the *severity* as the sum of the absolute SPEI values during the drought event; the *intensity* and the *Lowest SPEI* refer to the mean and lowest value of SPEI respectively during the drought event.

@@

### Statistical analysis 
The severe drought events since 1901 were identified using SPEI-12 and regional climatic data (Figure 2). They were characterized in terms of duration, severity, intensity [see @Spinoni2015]. In a first step we explored temporal trends of EVI and BAI variables. Temporal trends of $EVI_{mean}$ were examined at the pixel scale, using the Mann–Kendall nonparametric test. Temporal trends for BAI were assessed using mean BAI chronologies. The impact of drought in greenness and growth was exploring using the $EVI_{sa}$ and the mean RWI site chronologies. Additionally the relationships between climatic variables and tree-growth variables (RWI and BAI site chronologies) were assessed using bootstrapped Pearson's correlations estimated using treeclim` [@Zang2015]. The non-climatic disturbance impacts on tree-growth were evaluated using the site disturbance chronologies.

@@

## Results
### Time trends in vegetation greenness
The analysis of time trends in greenness showed that 78.9% of the EVI pixels followed a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the study period were recorded during the 2005 drought, and the minimum EVI values were expressed in the northern (dry) population (Figure 3a). A "browning" episode ($\mathrm{EVI_{sa}} < -1$) was found during this drought event, whereas no changes in greenness in response to the 2012 drought were detected (Figure 3b).

### Analysis of radial-growth trends and disturbances 
The trees of the southern population were older than those from the northern one. In addition, trees from the southern population at high elevation were taller and their growth was significantly greater than that of trees from the other two sites, despite the competition measured as plot basal area was greatest in CA-High (Table 1, Figure 4a). The growth and height of trees from the northern and the low-elevation southern population proved similar (Figures 4a and S3a). Only trees from the southern sites (*i.e.* the wetter exposure) showed significant positive growth trends since the late 1970s (Figure 4a), this trend being far more pronounced for the high elevation, the wetter, colder site (CA-High).

@@

The response of tree growth to water availability was greater than to temperatures. Cumulative precipitation of the hydrological year and seasonal SPEI values (*i.e.* for the Hydrological year, Spring and Summer) were the climatic variables exhibiting the highest (positive) relationship with growth for all populations (Figure S6a). Nevertheless, differences appeared between northern and southern populations: the positive relationship with SPEI was highest in the more xeric northern population (r > 0.6 *vs.* r < 0.5; Figure S6a).

The northern site (SJ) showed two major release events (GC > 50% occurring in more than 50% of trees sampled): the first during the 1940s (the most evident) and the second in 1995-2000 (Figure 4b). These periods alternated with periods of suppression. By contrast, the two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.

### Resilience to drought events at the ecosystem and individual-tree levels
Resilience and resistance varied in the same direction whereas recovery varied inversely to resilience and resistance. Resilience metrics of tree-growth for drought events since 1950 (*i.e.* shared period among the three chronologies excluding the juvenile years, Table S3) revealed a positive relationship between drought severity and recovery, significant for all oak populations (Figure 5a). A similar pattern was found for resilience but proved significant only for SJ. Importantly, non-significant patterns resulted when we excluded 1995, except for recovery in SJ (Figure S5). The trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area, particularly SJ where our results suggest a major release event also after 1995 (Figure 4b).

During the last two drought events, resilience metrics for greenness and tree growth significantly differed between drought events (Table S1). The 2005 drought event reduced greenness and growth more than that of 2012 (Table S2) but the metrics of resilience generally covaried in the same direction during those two years. For EVI, resilience and resistance values were significantly higher for 2012, the most severe event, than for 2005 (Table S2; Figure 5b); whereas recovery values were higher for 2005 than for the 2012 drought event. For BAI, the resilience, resistance and recovery values were higher for 2012 than for 2005 (Table S2, Figure 5c).

@@
